---
title: "DATA 607 dataset 3"
author: "Susanna Wong"
date: "2023-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load CSV into R

Load the csv file into R.

```{r, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(DT)
raw_data <- read.csv('https://raw.githubusercontent.com/suswong/DATA-607-Project-2/main/Gasoline_Retail_Prices_Weekly_Average_by_Region__Beginning_2007.csv')
raw_data$Date <- as.Date(raw_data$Date,
                        format = "%m/%d/%Y")
datatable(raw_data)
```

## Tidy the data
```{r}
newtable <- raw_data
colnames(newtable) <- c("Date", "New.York.State", "Albany","Batavia","Binghamton","Buffalo", "Dutchess", "Elmira,","Glens.Falls","Ithaca", "Kingston","Nassau","New.York.City","Rochester", "Syracuse","Utica","Watertown","White.Plains")

```

## From Wide Format to Long Format

Pivot the table so that each color is in a row.

```{r}
colnames(raw_data)
library(ggplot2)
long_table <- newtable %>% 
  pivot_longer(cols = !c('Date'),names_to = "region", values_to = "average_price" )
datatable(long_table)
```
```{r}
colnames(raw_data)
library(ggplot2)
long_table <- newtable %>% 
  pivot_longer(cols = !c('Date'), names_to = "region", values_to = "average_price" )
datatable(long_table)
```
## Data Analysis

### Create a line graph to view the trend or compare the prices of gasoline for each region.
Overall, we can see most region follows the same trend. Overall, there is a increase in average weekly prices  from mid 2007 to around 2012 for all region.  There is a decrease in average weekly prices from around 2012 to around 2017. The average weekly prices increase from 2017 to present. 
```{r}
library(ggplot2)
ggplot(long_table, aes(x = Date, y = average_price, colour = region)) +
  geom_smooth() + scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) 

```

# Compare the trend with the crude oil prices

# Annual US Crude Oil Prices from 1900

The US crude oil annual prices from 1900 can be found here: https://www.eia.gov/dnav/pet/hist/LeafHandler.ashx?n=pet&s=f000000__3&f=a

# Load CSV into R

The website provided a XLS file of the crude oil prices. The file was save as a csv file. 
```{r}
crude_oil_raw_data <- read.csv('https://raw.githubusercontent.com/suswong/DATA-607-Project-2/main/crude%20oil%20prices%20from%20eia.gov.csv')
```

# Tidy Data

Remove the first two rows and rename the columns. 
```{r}
colnames(crude_oil_raw_data) <- c("Year", "Price")
crude_oil <- crude_oil_raw_data
tidy_crude_oil <- crude_oil [-1,]
tidy_crude_oil <- tidy_crude_oil [-1,]
```

# Filter out crude oil prices prior to 2007
```{r}
filtered_crude_oil <- tidy_crude_oil[as.numeric(tidy_crude_oil$Year) >2006,] 
datatable(filtered_crude_oil)

# filtered_crude_oil <- tidy_crude_oil %>% 
#   filter(Year >2006)
# datatable(filtered_crude_oil)
```

# Plot the crude oil

The annual US crude oil prices have similar trends as the average weekly gasoline prices in NY.
```{r}
ggplot(filtered_crude_oil, aes(x = Year, y = Price, group=1)) +
  geom_line()+ ggtitle("Annual Crude Oil Price from 2007 to 2022") + 
  xlab("Year") + 
  ylab("US Crude Oil Price (Dollars/Barrel)") 
ggplot(long_table, aes(x = Date, y = average_price, colour = region)) +
  geom_line() + scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) + ggtitle("Average Weekly Gasoline Prices in Different NY Region from 2007 to Present") + 
  ylab("Average Weekly Gasoline Prices") + 
  xlab("Year") 
```


